When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-27 doc/paper/body.tex
kidn of done with methods, need to decide where to put paragraph on soil moisture

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with
temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$),
$R_{net}$ is the net radiation minus ground heat flux, $g_a$ is
aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific
heat of air at constant pressure, $\gamma$ is the psychometric
constant, and $g_s$ is the stomatal conductance (Table
\ref{definitions}).

@@

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; A}$). So,
$g_{1-leaf}$ is a leaf-scale term reflecting the trade-off of water for
carbon uptake. The higher $g_{1-leaf}$, the more open the stomata and
the more they release water in exchange for carbon.

@@

While Eq. (\ref{medlyn}) can be used in PM (Eq. (1)), it will make
analytical work with the function intractable because $GPP$ is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Eq. (\ref{medlyn}). Therefore, in order to
derive the response of ET to VPD, we must account for the functional
relationship between GPP, ET, and VPD, and its effect on stomatal
conductance. We can use aforementioned semi-empirical results of
\citet{Zhou_2015} as a tool to approach this
problem. \citet{Zhou_2015}, showed that  underlying Water Use
Efficiency (uWUE):

@@

  is relatively constant across time and moisture conditions within a
  plant functional type, and correctly captures a constant
  relationship between GPP, ET and VPD over a diurnal cycle during the
  growing season \citep{Zhou_2014}. The theoretical derivation of the
  square root VPD dependence in $uWUE$ leverages the same assumptions
  used in \cite{MEDLYN_2011} to derive the square-root VPD dependence
  of the stomatal conductance model (Eq.  (\ref{medlyn})).  We can use
  uWUE to remove the $GPP$ dependence from $g_s$ in a way that makes
  PM analytically tractable:

@@

By accounting for photosynthesis changes in ecosystem conductance, with
Eq. (\ref{et}) we have derived for the first time, using recent
results \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017}, ET
explicitly as function of environmental variables and two
plant-specific constants, the slope parameter ($g_1$), and uWUE, both
reflecting water conservation strategy. The slope parameter is related
to the willingness of stomata to trade water for CO$_2$ and to keep
stomata open. uWUE is a semi-empirical ecosystem-scale constant
related to how WUE changes with VPD (specifically $VPD^{-1/2}$). It is
also roughly proportional to physical constants:

@@

With Equation \ref{et} we can take the partial derivative with respect
to VPD:

@@

  deriving an analytical framework for ecosystem response to
  atmospheric demand perturbations with environmental conditions held
  fixed. There are a few subtleties to taking the derivative in
  Eq. (\ref{d_et}): $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are
  functionally related, so while taking the derivative we evaluate
  $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
  {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
  \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
  \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
      fixed}}$. $RH$ and $e_s$ are assumed to be approximately
  independent, which is supported by data (not shown).

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation (E) contributions to ET remain small
  relative to transpiration (T). This should be more true during the
  growing season. The ratio of E to T may increase immediately after
  rainfall events due to high soil moisture and ponding, but VPD is
  generally low anyways during these times. However, some plant types
  allow for systematically larger contributions of E in ET,
  particularly those with sparse canopies and smaller relative amounts
  of transpiration. We therefore might expect that the theory will be
  most applicable to forest PFTs, which will be most strongly coupled
  to the boundary layer due to larger surface roughness, and will also
  generally have the highest ratios of transpiration to
  evaporation. Finally, we assume that $g_1$ and $uWUE$ are constant
  with respect to the conceptual VPD perturbation. Both quantities
  have been shown to be relatively constant with respect to changes in
  % TODO: make this a general discussion on partial derivatives and
  % what things are constant/how to interpret?
  VPD \citep{Franks_2017, Zhou_2014}. These parameters will however
  vary with plant species and characteristics \citep[e.g. wood density,
  ][]{Lin_2015}, as well as environmental conditions including soil
  water content and temperature.
  % TODO: write exact parameter ranges we use, and introduce
  % scaling/sign term here?
  % \item
  To account for both this variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

>>>>

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with
temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$),
$R_{net}$ is the net radiation minus ground heat flux, $g_a$ is
aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific
heat of air at constant pressure, $\gamma$ is the psychometric
constant, and $g_s$ is the stomatal conductance (Table
\ref{definitions}). The issue with PM is that $g_s$ is a function of
carbon uptake, which is has a strong functionally relation to ET
through stomatal function. Therefore, when PM is formulated in terms
of $g_s$, it is an implicit funciton of ET as a function of itself
rather than an explicit function of ET. Here we will derive a new form
of PM in which ET is an explicit function of plant parameters and
environmental conditions, and use it to assess the ecosystem scale
response to VPD (by taking a partial derivative).

@@

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; A}$)
\citep{Farquhar_1980, Katul_2009}. So,
$g_{1-leaf}$ is a leaf-scale term reflecting the trade-off of water for
carbon uptake. The higher $g_{1-leaf}$, the more open the stomata and
the more they release water in exchange for carbon.

@@

  While Eq. (\ref{medlyn}) can be used in PM (Eq. (1)), it will make
  analytical work with the function intractable because $GPP$ is
  functionally related to ET itself. Additionally, a perturbation to
  VPD should induce a physiological plant response that will alter GPP
  and cause an indirect change in stomatal conductance, in addition to
  the direct effect of VPD in Eq. (\ref{medlyn}). Therefore, in order
  to derive the response of ET to VPD, we must account for the
  functional relationship between GPP, ET, and VPD, and its effect on
  stomatal conductance. We can use aforementioned semi-empirical
  results of \citet{Zhou_2015}, which were inspired optimal
  photosynthesis theory, as a tool to approach this
  problem. \citet{Zhou_2015}, showed that underlying Water Use
  Efficiency (uWUE):

@@

  is relatively constant across time within a plant functional type,
  and correctly captures a constant relationship between GPP, ET and
  VPD over a diurnal cycle during the growing season
  \citep{Zhou_2014}. The theoretical derivation of the square root VPD
  dependence in $uWUE$ leverages the same assumptions used in
  \cite{MEDLYN_2011} to derive the square-root VPD dependence of the
  stomatal conductance model (Eq.  (\ref{medlyn})).  We can use uWUE
  to remove the $GPP$ dependence from $g_s$ in a way that makes PM
  analytically tractable:

@@

  By accounting for photosynthesis changes in ecosystem conductance,
  with Eq. (\ref{et}) we have used recent results
  \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} to derive
  ET explicitly as function of environmental variables and two
  plant-specific constants, the slope parameter ($g_1$), and uWUE. For
  the first time we have removed the implicit dependence of ET on
  itself through the stomatal conductance term, and we have also
  replaced the added complexity of a stomatal conductance reduction
  factor and a photosythesis model with a signle paramter (uWUE). Both
  the plant paramters reflect water conservation strategy.The
  slope parameter is related to the willingness of stomata to trade
  water for CO$_2$ and to keep stomata open (carbon cost in terhms of
  water)). uWUE is a semi-empirical ecosystem-scale constant related
  to how WUE changes with VPD (specifically $VPD^{-1/2}$). It is also
  roughly proportional to physical constants:

@@

With Equation \ref{et} we can take the partial derivative of ET with respect
to VPD to understand whether VPD drives or reduces ecosystem-scale ET:

@@

  deriving an analytical framework for ecosystem response to
  atmospheric demand with environmental conditions held
  fixed. There are a few subtleties to taking the derivative in
  Eq. (\ref{d_et}): $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are
  functionally related, so while taking the derivative we evaluate
  $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
  {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
  \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
  \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
      fixed}}$. $RH$ and $e_s$ are assumed to be approximately
  independent, which is supported by data (not shown).

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation (E) contributions to ET remain small
  relative to transpiration (T). Again, this should be more true
  during the growing season. The ratio of E to T may increase
  immediately after rainfall events due to high soil moisture and
  ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of E in ET, particularly those with sparse canopies
  and smaller relative amounts of transpiration. We therefore might
  expect that the theory will be most applicable to forest PFTs, which
  will be most strongly coupled to the boundary layer due to larger
  surface roughness, and will also generally have the highest ratios
  of transpiration to evaporation. Finally, we assume that $g_1$ and
  $uWUE$ are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in
  % TODO: make this a general discussion on partial derivatives and
  % what things are constant/how to interpret?
  VPD \citep{Franks_2017, Zhou_2014}. These parameters will however
  vary with plant species and characteristics \citep[e.g. wood density,
  ][]{Lin_2015}, as well as environmental conditions including soil
  water content and temperature.
  % TODO: write exact parameter ranges we use, and introduce
  % scaling/sign term here?
  % \item
  To account for both this variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

  ** comment on temperature and acclimation??? **
  ** impact of T is much smaller than VPD? **
  ** paragraph on soil moisture **
